Womb for One More? Fertility Expansion and the Limits of Early Health Investment

### 2025-07-27 docs/paper.Rmd
fix math equation formatting in empirical strategy
Corrected the display of the difference-in-differences equation in both the R Markdown and LaTeX source to use proper math environments.
revise paper title and introduction for clarity
Updated the paper's title, subtitle, and abstract to focus on early-life health outcomes and the impact of Manila's 2000 contraceptive ban. The introduction was rewritten to emphasize health outcomes (birth weight, perceived birth size, vaccinations) rather than nutrition alone, clarify the empirical strategy, and provide more context on the policy and its quasi

---
title: "Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility affects early-life nutritional outcomes in low-income urban households. I exploit a natural experiment created by a 1990 policy in Manila, Philippines, which banned modern contraceptives from city-run health facilities. Using a difference-in-differences framework and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the reduced-form impact of the policy on child height-for-age and weight-for-height. [Will add more after data analysis]
output: 
  rticles::aea_article:
    keep_tex: true
    latex_engine: pdflatex
bibliography: references.bib
---

@@

The trade-off between child quantity and child quality is a foundational concept in the economics of the family. First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{becker1973interaction, becker1976child}, this framework posits that parents allocate finite resources—both financial and non-financial—across children. An increase in fertility reduces the resources available per child and, under binding constraints, may lead to lower investments in health, education, and other forms of human capital. This mechanism has served as an explanatory model for changes in fertility behavior and the evolution of population structures in low- and middle-income countries.

Empirical investigations of the quantity–quality trade-off have focused primarily on educational outcomes. Studies in both high-income and low-income settings have examined the effects of fertility on school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality. Other outcomes, such as nutritional status, are early-onset or biologically constrained. They also tend to be less responsive to remediation later in life. A child who suffers from chronic malnutrition may exhibit permanently reduced cognitive capacity and face limits in physical development that affect long-run productivity regardless of subsequent educational access \citep{hoddinott2013adult, grantham2007development}.

The exclusion of nutritional outcomes from much of the empirical literature leaves an important dimension of the trade-off untested. Nutritional investments in early childhood are essential to early childhood development and long-term outcomes \citep{victora2008maternal, hoddinott2013adult}. They shape brain development, immune system functioning, and physical stature, and they have been shown to predict later-life earnings and health outcomes across a wide range of settings \citep{grantham2007development, alderman2006long}. The biological irreversibility of early-life nutritional deficits further distinguishes them from other forms of investment. Educational deficits may be partially remediable; nutritional failures often are not. A credible estimate of the trade-off between fertility and child quality must account for nutrition if it aims to assess the full set of consequences associated with fertility shocks.

This study addresses this gap by examining the nutritional effects of a localized, exogenous increase in fertility in the Philippines. In 1990, the mayor of Manila implemented an executive policy that prohibited the provision of modern contraceptives in all city-run health facilities. The order removed access to pills, condoms, intrauterine devices, and related public health materials and instructed healthcare providers to offer only natural family planning methods. This policy remained in place for nearly a decade and affected only the jurisdiction of the Manila city government. The national government did not implement a comparable restriction, and surrounding cities within Metro Manila continued to provide access to modern contraceptives. The policy thus created a spatial and temporal discontinuity in contraceptive access that was uncorrelated with underlying fertility preferences or concurrent shifts in household income or governance. As a result, the Manila ban serves as a quasi-experimental source of variation in fertility exposure among poor urban households.

I use this natural experiment to estimate the causal effect of increased fertility on child nutrition. The analysis relies on nationally representative data from multiple waves of the Philippine Demographic and Health Survey (DHS), which provide data on household structure and maternal characteristics, as well as measurements of child anthropometry. The outcomes of interest are height-for-age and weight-for-height z-scores, which serve as standardized indicators of chronic and acute malnutrition, respectively. These outcomes are widely used in the global health and development literature and capture nutritional deprivation over both long and short time horizons \citep{victora2008maternal}. The empirical strategy follows a difference-in-differences design that compares child outcomes in Manila and comparable urban areas before and after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the absence of the contraceptive ban, nutritional trends in Manila would have evolved in parallel with those in comparison cities. Second, any other policy or economic shocks affecting Manila during the study period must not coincide precisely with the timing and scope of the contraceptive policy. I test these assumptions using falsification checks, placebo comparisons, and robustness specifications that include city-specific time trends, maternal fixed effects, and controls for baseline demographic differences.

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. I then estimate reduced-form effects of policy exposure on nutritional outcomes for children under five years of age. Finally, I examine heterogeneity in effects across subsamples defined by maternal education, household wealth, and access to prenatal care. These dimensions serve as proxies for household resource availability and capacity to buffer the nutritional consequences of fertility increases.

This study contributes to the literature in several important ways. It provides new evidence on how increases in fertility—caused by policy restrictions on family planning—can affect child nutrition in poor, urban communities. Most past research has focused on education, but this study expands the idea of child quality to include biological measures such as stunting and wasting. It also adds to the small number of studies that use unexpected changes in reproductive health policy to examine long-term effects on children’s well-being. More broadly, the results show that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public services.

@@

\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict}

>>>>

---
title: "Womb for One More? Fertility Expansion and the Limits of Early Health Investment"
subtitile: "Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines"
short: "The Q-Q Trade-off in the Philippine Context"
journal: "" # AER, AEJ, PP, JEL
month: "`r format(Sys.Date(), '%m')`"
year: "`r format(Sys.Date(), '%Y')`"
jel:
  - J13  # Fertility; Childbearing; Family Planning; Population Policy
  - I15  # Health and Economic Development
  - O15  # Human Resources; Human Development; Income Distribution; Migration
keywords:
  - fertility shocks
  - child nutrition
  - quantity-quality trade-off
  - urban poverty
  - Philippines
author:
  - name: Erika Salvador
    firstname: Erika
    surname: Salvador
    email: esalvador28@amherst.edu
    affiliation: Amherst College
  - name: Caroline Theoharides
    firstname: Caroline
    surname: Theoharides
    email: ctheoharides@amherst.edu
    affiliation: Amherst College
acknowledgements: |
  This research was made possible by the Schupf Fellowship at Amherst College. I am indebted to my faculty adviser, Professor Caroline Theoharides, for her mentorship and support throughout the development of this project. I also wish to thank Faculty Director Professor Amelie Hastie and the campus partners whose institutional support enabled this work. I am further grateful to the Departments of Economics and Mathematics & Statistics for their academic support. The views expressed and any errors contained herein are entirely my own responsibility.
abstract: |
  This paper examines whether increased fertility, driven by reduced access to modern contraception, affects early-life health outcomes in low-income urban households. I exploit a natural experiment created by a 2000 policy in Manila, Philippines—Executive Order No. 003—which effectively banned the provision of modern contraceptives in all city-run health facilities. Using a difference-in-differences design and nationally representative data from the Philippine Demographic and Health Surveys, I estimate the policy’s impact on child health by comparing outcomes in Manila before and after the ban to those in comparable urban areas unaffected by the policy. The outcomes of interest include birth weight (in grams), the mother’s perceived size of the child at birth, and receipt of key childhood vaccinations.
output: 
  rticles::aea_article:
    keep_tex: true
    latex_engine: pdflatex
bibliography: references.bib
---

@@

The trade-off between child quantity and child quality is a foundational concept in the economics of the family. First articulated by \citet{becker1960economic} and extended in subsequent models of household behavior \citep{becker1973interaction, becker1976child}, this framework posits that parents allocate finite resources—both financial and non-financial—across children. An increase in fertility dilutes the resources available per child and, under binding constraints, may lead to reduced investments in each child’s human capital.

Empirical investigations of the quantity-quality trade-off have focused primarily on educational outcomes. Most studies examine how increases in maternal fertility—or equivalently, the number of siblings a child has—affect school enrollment, grade progression, test scores, and completed years of schooling \citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These outcomes serve as accessible proxies for long-run human capital accumulation, but they represent only one dimension of child quality. Other dimensions of child quality–such as health at birth and early childhood health investments—are early-onset or biologically constrained, and they tend to be less amenable to remediation later in life. For instance, babies born with very low birth weight face higher risks of infant mortality and long-term developmental deficits that cannot be fully reversed \citep{cook2015understanding}. Similarly, a child who misses critical early vaccinations is more vulnerable to infectious diseases that can kill or disable them, which results to lifelong negative health impacts \citep{costa2024child}. Such early-life health setbacks may permanently impair cognitive development and physical growth. This limits long-run productivity regardless of subsequent improvements in schooling or incom

The omission of these health outcomes from much of the empirical literature leaves important aspects of the quantity-quality trade-off unexamined. Investments in maternal and child health—such as adequate prenatal care leading to healthy birth weight and timely childhood immunizations—are important to early development and long-term outcomes \citep{victora2008maternal, hoddinott2013adult}. They shape brain development, immune system functioning, and physical stature, and they have been shown to predict later-life earnings and health outcomes across a wide range of settings \citep{grantham2007development, alderman2006long}. Unlike educational gaps, which may be partially addressed later in life, early health deficits—such as poor nutrition or missed vaccinations—are often irreversible \citep{victora2008maternal, hoddinott2013adult}. As a result, any credible evaluation of the quantity-quality trade-off must include early-life health outcomes to fully capture the long-term consequences of different fertility shocks.

This study addresses that gap by examining how a localized, exogenous increase in fertility influenced early-life health outcomes in the Philippines. In 2000, then–Mayor José “Lito” Atienza of Manila City issued Executive Order No. 003, which effectively barred the provision and promotion of modern contraceptives in all city-run health facilities. The directive eliminated access to pills, condoms, intrauterine devices, and other forms of modern family planning, and mandated that public healthcare providers offer only natural methods. The policy remained in effect for nearly a decade and applied solely within the jurisdiction of the Manila city government. No comparable restriction was implemented at the national level, and neighboring cities within Metro Manila continued to provide access to modern contraceptives. This policy created a clear spatial and temporal discontinuity in access to family planning services—one that was plausibly unrelated to changes in fertility preferences, household income, or broader governance trends. As such, the Manila contraceptive ban provides a quasi-experimental shock to fertility exposure among poor urban households.

I use this natural experiment to estimate the causal impact of increased family size on early-childhood health outcomes. The analysis uses nationally representative data from multiple rounds of the Philippine Demographic and Health Survey (DHS), which include detailed information on household structure, maternal characteristics, and child health indicators. The primary outcomes are birth weight (measured in grams, when reported) and the mother’s subjective assessment of the child’s size at birth—both of which serve as indicators of neonatal health status. In addition, I assess whether the child received key early-life vaccinations, including those against polio, diphtheria–pertussis–tetanus (DPT), and measles, as a measure of essential postnatal health investment. These outcomes are widely recognized indicators of child well-being and reflect both biological and behavioral dimensions of parental investment in early development. The empirical strategy follows a difference-in-differences design that compares child outcomes in Manila and comparable urban areas before and after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the absence of the contraceptive ban, trends in child health outcomes in Manila would have evolved in parallel with those in the comparison cities. Second, no other policy changes or economic shocks in Manila during the study period coincided with the timing and scope of the contraceptive policy. I probe these assumptions using falsification tests, placebo comparisons, and robustness checks that include city-specific time trends, mother fixed effects, and controls for baseline demographic differences.

The analysis proceeds in three stages. I first replicate existing work \citep{dumas2019sex} to confirm that the contraceptive ban led to an increase in fertility among affected women. Next, I estimate the reduced-form effects of exposure to the policy on the selected child health outcomes (birth weight, perceived birth size, and vaccination uptake). Finally, I examine heterogeneity in these effects across subgroups defined by maternal education, household wealth, and access to prenatal care. These characteristics proxy for a household’s resources and its capacity to buffer the impact of increased family size on child health.

This study contributes to the literature in several important ways. It provides new evidence on how an externally induced increase in family size, resulting from a local restriction on contraceptive access,  can affect child health in poor urban communities. Whereas most past research on the quantity-quality trade-off has focused on education, this study broadens the notion of child “quality” to include early-life health measures such as birth weight and immunization status. It also adds to the small number of studies that leverage unexpected changes in reproductive health policy to examine long-term effects on children’s well-being. More broadly, the findings indicate that local restrictions on family planning can unintentionally harm children’s health, especially in settings where families already face poverty, food insecurity, and limited public health services.

@@

$$ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} $$

### 2025-07-27 docs/paper.tex
fix math equation formatting in empirical strategy
Corrected the display of the difference-in-differences equation in both the R Markdown and LaTeX source to use proper math environments.
revise paper title and introduction for clarity
Updated the paper's title, subtitle, and abstract to focus on early-life health outcomes and the impact of Manila's 2000 contraceptive ban. The introduction was rewritten to emphasize health outcomes (birth weight, perceived birth size, vaccinations) rather than nutrition alone, clarify the empirical strategy, and provide more context on the policy and its quasi

\title{Nutritional Deficits and the Quantity-Quality Trade-off: Evidence
from an Exogenous Fertility Shock in Low-Income Urban Settings in the
Philippines}
\shortTitle{The Q-Q Trade-off in the Philippine Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

@@

\begin{abstract}
This paper examines whether increased fertility affects early-life
nutritional outcomes in low-income urban households. I exploit a natural
experiment created by a 1990 policy in Manila, Philippines, which banned
modern contraceptives from city-run health facilities. Using a
difference-in-differences framework and nationally representative data
from the Philippine Demographic and Health Surveys, I estimate the
reduced-form impact of the policy on child height-for-age and
weight-for-height. {[}Will add more after data analysis{]}
\end{abstract}

@@

The trade-off between child quantity and child quality is a foundational
concept in the economics of the family. First articulated by
\citet{becker1960economic} and extended in subsequent models of
household behavior \citep{becker1973interaction, becker1976child}, this
framework posits that parents allocate finite resources---both financial
and non-financial---across children. An increase in fertility reduces
the resources available per child and, under binding constraints, may
lead to lower investments in health, education, and other forms of human
capital. This mechanism has served as an explanatory model for changes
in fertility behavior and the evolution of population structures in low-
and middle-income countries.

Empirical investigations of the quantity--quality trade-off have focused
primarily on educational outcomes. Studies in both high-income and
low-income settings have examined the effects of fertility on school
enrollment, grade progression, test scores, and completed years of
schooling
\citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
outcomes serve as accessible proxies for long-run human capital
accumulation, but they represent only one dimension of child quality.
Other outcomes, such as nutritional status, are early-onset or
biologically constrained. They also tend to be less responsive to
remediation later in life. A child who suffers from chronic malnutrition
may exhibit permanently reduced cognitive capacity and face limits in
physical development that affect long-run productivity regardless of
subsequent educational access
\citep{hoddinott2013adult, grantham2007development}.

The exclusion of nutritional outcomes from much of the empirical
literature leaves an important dimension of the trade-off untested.
Nutritional investments in early childhood are essential to early
childhood development and long-term outcomes
\citep{victora2008maternal, hoddinott2013adult}. They shape brain
development, immune system functioning, and physical stature, and they
have been shown to predict later-life earnings and health outcomes
across a wide range of settings
\citep{grantham2007development, alderman2006long}. The biological
irreversibility of early-life nutritional deficits further distinguishes
them from other forms of investment. Educational deficits may be
partially remediable; nutritional failures often are not. A credible
estimate of the trade-off between fertility and child quality must
account for nutrition if it aims to assess the full set of consequences
associated with fertility shocks.

This study addresses this gap by examining the nutritional effects of a
localized, exogenous increase in fertility in the Philippines. In 1990,
the mayor of Manila implemented an executive policy that prohibited the
provision of modern contraceptives in all city-run health facilities.
The order removed access to pills, condoms, intrauterine devices, and
related public health materials and instructed healthcare providers to
offer only natural family planning methods. This policy remained in
place for nearly a decade and affected only the jurisdiction of the
Manila city government. The national government did not implement a
comparable restriction, and surrounding cities within Metro Manila
continued to provide access to modern contraceptives. The policy thus
created a spatial and temporal discontinuity in contraceptive access
that was uncorrelated with underlying fertility preferences or
concurrent shifts in household income or governance. As a result, the
Manila ban serves as a quasi-experimental source of variation in
fertility exposure among poor urban households.

I use this natural experiment to estimate the causal effect of increased
fertility on child nutrition. The analysis relies on nationally
representative data from multiple waves of the Philippine Demographic
and Health Survey (DHS), which provide data on household structure and
maternal characteristics, as well as measurements of child
anthropometry. The outcomes of interest are height-for-age and
weight-for-height z-scores, which serve as standardized indicators of
chronic and acute malnutrition, respectively. These outcomes are widely
used in the global health and development literature and capture
nutritional deprivation over both long and short time horizons
\citep{victora2008maternal}. The empirical strategy follows a
difference-in-differences design that compares child outcomes in Manila
and comparable urban areas before and after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the
absence of the contraceptive ban, nutritional trends in Manila would
have evolved in parallel with those in comparison cities. Second, any
other policy or economic shocks affecting Manila during the study period
must not coincide precisely with the timing and scope of the
contraceptive policy. I test these assumptions using falsification
checks, placebo comparisons, and robustness specifications that include
city-specific time trends, maternal fixed effects, and controls for
baseline demographic differences.

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2019sex} to confirm that the contraceptive ban led to an
increase in fertility among affected women. I then estimate reduced-form
effects of policy exposure on nutritional outcomes for children under
five years of age. Finally, I examine heterogeneity in effects across
subsamples defined by maternal education, household wealth, and access
to prenatal care. These dimensions serve as proxies for household
resource availability and capacity to buffer the nutritional
consequences of fertility increases.

This study contributes to the literature in several important ways. It
provides new evidence on how increases in fertility---caused by policy
restrictions on family planning---can affect child nutrition in poor,
urban communities. Most past research has focused on education, but this
study expands the idea of child quality to include biological measures
such as stunting and wasting. It also adds to the small number of
studies that use unexpected changes in reproductive health policy to
examine long-term effects on children's well-being. More broadly, the
results show that local restrictions on family planning can
unintentionally harm children's health, especially in settings where
families already face poverty, food insecurity, and limited public
services.

@@

\text{Nutrition}\_\{ict\} = \alpha + \beta \cdot (\text{Post}\_t
\times \text{Manila}\emph{c) + X}\{ict\}'\gamma + \lambda\_c +
\tau\emph{t + \varepsilon}\{ict\}

>>>>

\title{Womb for One More? Fertility Expansion and the Limits of Early
Health Investment}
\shortTitle{The Q-Q Trade-off in the Philippine Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

@@

\begin{abstract}
This paper examines whether increased fertility, driven by reduced
access to modern contraception, affects early-life health outcomes in
low-income urban households. I exploit a natural experiment created by a
2000 policy in Manila, Philippines---Executive Order No.~003---which
effectively banned the provision of modern contraceptives in all
city-run health facilities. Using a difference-in-differences design and
nationally representative data from the Philippine Demographic and
Health Surveys, I estimate the policy's impact on child health by
comparing outcomes in Manila before and after the ban to those in
comparable urban areas unaffected by the policy. The outcomes of
interest include birth weight (in grams), the mother's perceived size of
the child at birth, and receipt of key childhood vaccinations.
\end{abstract}

@@

The trade-off between child quantity and child quality is a foundational
concept in the economics of the family. First articulated by
\citet{becker1960economic} and extended in subsequent models of
household behavior \citep{becker1973interaction, becker1976child}, this
framework posits that parents allocate finite resources---both financial
and non-financial---across children. An increase in fertility dilutes
the resources available per child and, under binding constraints, may
lead to reduced investments in each child's human capital.

Empirical investigations of the quantity-quality trade-off have focused
primarily on educational outcomes. Most studies examine how increases in
maternal fertility---or equivalently, the number of siblings a child
has---affect school enrollment, grade progression, test scores, and
completed years of schooling
\citep{rosenzweig1980testing, black2005more, angrist2010multiple}. These
outcomes serve as accessible proxies for long-run human capital
accumulation, but they represent only one dimension of child quality.
Other dimensions of child quality--such as health at birth and early
childhood health investments---are early-onset or biologically
constrained, and they tend to be less amenable to remediation later in
life. For instance, babies born with very low birth weight face higher
risks of infant mortality and long-term developmental deficits that
cannot be fully reversed \citep{cook2015understanding}. Similarly, a
child who misses critical early vaccinations is more vulnerable to
infectious diseases that can kill or disable them, which results to
lifelong negative health impacts \citep{costa2024child}. Such early-life
health setbacks may permanently impair cognitive development and
physical growth. This limits long-run productivity regardless of
subsequent improvements in schooling or incom

The omission of these health outcomes from much of the empirical
literature leaves important aspects of the quantity-quality trade-off
unexamined. Investments in maternal and child health---such as adequate
prenatal care leading to healthy birth weight and timely childhood
immunizations---are important to early development and long-term
outcomes \citep{victora2008maternal, hoddinott2013adult}. They shape
brain development, immune system functioning, and physical stature, and
they have been shown to predict later-life earnings and health outcomes
across a wide range of settings
\citep{grantham2007development, alderman2006long}. Unlike educational
gaps, which may be partially addressed later in life, early health
deficits---such as poor nutrition or missed vaccinations---are often
irreversible \citep{victora2008maternal, hoddinott2013adult}. As a
result, any credible evaluation of the quantity-quality trade-off must
include early-life health outcomes to fully capture the long-term
consequences of different fertility shocks.

This study addresses that gap by examining how a localized, exogenous
increase in fertility influenced early-life health outcomes in the
Philippines. In 2000, then--Mayor José ``Lito'' Atienza of Manila City
issued Executive Order No.~003, which effectively barred the provision
and promotion of modern contraceptives in all city-run health
facilities. The directive eliminated access to pills, condoms,
intrauterine devices, and other forms of modern family planning, and
mandated that public healthcare providers offer only natural methods.
The policy remained in effect for nearly a decade and applied solely
within the jurisdiction of the Manila city government. No comparable
restriction was implemented at the national level, and neighboring
cities within Metro Manila continued to provide access to modern
contraceptives. This policy created a clear spatial and temporal
discontinuity in access to family planning services---one that was
plausibly unrelated to changes in fertility preferences, household
income, or broader governance trends. As such, the Manila contraceptive
ban provides a quasi-experimental shock to fertility exposure among poor
urban households.

I use this natural experiment to estimate the causal impact of increased
family size on early-childhood health outcomes. The analysis uses
nationally representative data from multiple rounds of the Philippine
Demographic and Health Survey (DHS), which include detailed information
on household structure, maternal characteristics, and child health
indicators. The primary outcomes are birth weight (measured in grams,
when reported) and the mother's subjective assessment of the child's
size at birth---both of which serve as indicators of neonatal health
status. In addition, I assess whether the child received key early-life
vaccinations, including those against polio,
diphtheria--pertussis--tetanus (DPT), and measles, as a measure of
essential postnatal health investment. These outcomes are widely
recognized indicators of child well-being and reflect both biological
and behavioral dimensions of parental investment in early development.
The empirical strategy follows a difference-in-differences design that
compares child outcomes in Manila and comparable urban areas before and
after the onset of the policy.

The identification strategy rests on two key assumptions. First, in the
absence of the contraceptive ban, trends in child health outcomes in
Manila would have evolved in parallel with those in the comparison
cities. Second, no other policy changes or economic shocks in Manila
during the study period coincided with the timing and scope of the
contraceptive policy. I probe these assumptions using falsification
tests, placebo comparisons, and robustness checks that include
city-specific time trends, mother fixed effects, and controls for
baseline demographic differences.

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2019sex} to confirm that the contraceptive ban led to an
increase in fertility among affected women. Next, I estimate the
reduced-form effects of exposure to the policy on the selected child
health outcomes (birth weight, perceived birth size, and vaccination
uptake). Finally, I examine heterogeneity in these effects across
subgroups defined by maternal education, household wealth, and access to
prenatal care. These characteristics proxy for a household's resources
and its capacity to buffer the impact of increased family size on child
health.

This study contributes to the literature in several important ways. It
provides new evidence on how an externally induced increase in family
size, resulting from a local restriction on contraceptive access, can
affect child health in poor urban communities. Whereas most past
research on the quantity-quality trade-off has focused on education,
this study broadens the notion of child ``quality'' to include
early-life health measures such as birth weight and immunization status.
It also adds to the small number of studies that leverage unexpected
changes in reproductive health policy to examine long-term effects on
children's well-being. More broadly, the findings indicate that local
restrictions on family planning can unintentionally harm children's
health, especially in settings where families already face poverty, food
insecurity, and limited public health services.

@@

\[ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} \]
